---
title: "Twin Cities Rental Price Analysis"
author: "Ha Phan"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#### Load Libraries
```{r warning=FALSE, message=FALSE}
library(DBI)
library(RSQLite)
library(ggplot2)
library(dplyr)
library(corrplot)
library(psych)
library(knitr)
```

#### Connect to Database
```{r}
conn <- dbConnect(SQLite(), 
  "/Users/crystalphan/Desktop/Apartent rental project/twincities_rental.db")

df <- dbGetQuery(conn, "SELECT * FROM annual_summary")
cat("Rows loaded:", nrow(df), "\n")
cat("Columns:", ncol(df), "\n")
```

#### Preview Data
```{r}
head(df)
```

#### Descriptive Statistics
```{r}
describe(df[, c("avg_rent", "rent_studio", "rent_1br", 
                "rent_2br", "rent_3br", "rent_4br")])
```

#### Chart 1: Average Rent by Year (Twin Cities Trend)
```{r}
# Calculate average rent per year across all zip codes
yearly <- df %>%
  group_by(year) %>%
  summarise(avg_rent = mean(avg_rent))

# Plot the trend line
ggplot(yearly, aes(x = year, y = avg_rent)) +
  geom_line(color = "#185FA5", size = 1.2) +
  geom_point(color = "#185FA5", size = 3) +
  labs(
    title = "Average Rent Trend — Twin Cities (2015–2026)",
    x = "Year",
    y = "Average Rent ($)"
  ) +
  theme_minimal()
```

#### Chart 2: Rent Distribution by Zip Code (2025)
```{r}
# Filter to 2025 only
df_2025 <- df %>% filter(year == 2025)

# Histogram of rent distribution
ggplot(df_2025, aes(x = avg_rent)) +
  geom_histogram(bins = 20, fill = "#185FA5", color = "white") +
  labs(
    title = "Distribution of Average Rent by Zip Code (2025)",
    x = "Average Rent ($)",
    y = "Number of Zip Codes"
  ) +
  theme_minimal()
```

#### Chart 3: Top 10 Most Expensive vs Affordable Zip Codes (2025)
```{r fig.height=6}
# Get top 5 most expensive and top 5 most affordable
top5    <- df_2025 %>% arrange(desc(avg_rent)) %>% head(5)
bottom5 <- df_2025 %>% arrange(avg_rent) %>% head(5)
combined <- rbind(top5, bottom5)

ggplot(combined, aes(x = reorder(regionname, avg_rent), 
                      y = avg_rent, fill = avg_rent)) +
  geom_bar(stat = "identity") +
  scale_fill_gradient(low = "#B5D4F4", high = "#0C447C") +
  coord_flip() +
  labs(
    title = "Most Expensive vs Most Affordable Zip Codes (2025)",
    x = "Zip Code",
    y = "Average Rent ($)",
    fill = "Rent ($)"
  ) +
  theme_minimal()
```

#### Chart 4: Rent by Bedroom Size (HUD Fair Market Rents)
```{r}
# Average HUD rent by bedroom size across all zip codes
bedroom_avg <- df %>%
  summarise(
    Studio   = mean(rent_studio),
    `1 BR`   = mean(rent_1br),
    `2 BR`   = mean(rent_2br),
    `3 BR`   = mean(rent_3br),
    `4 BR`   = mean(rent_4br)
  ) %>%
  tidyr::pivot_longer(everything(), 
                       names_to = "bedroom", 
                       values_to = "avg_rent")

ggplot(bedroom_avg, aes(x = bedroom, y = avg_rent, fill = bedroom)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Average Fair Market Rent by Bedroom Size — Minneapolis MSA",
    x = "Bedroom Size",
    y = "Average Rent ($)"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

#### Correlation Matrix
```{r}
# Select numeric columns for correlation
cor_data <- df[, c("avg_rent", "rent_studio", "rent_1br", 
                    "rent_2br", "rent_3br", "rent_4br")]

# Calculate correlation matrix
cor_matrix <- cor(cor_data, use = "complete.obs")

# Plot it
corrplot(cor_matrix, 
         method = "color",
         type = "upper",
         tl.col = "black",
         tl.srt = 45,
         addCoef.col = "black",
         number.cex = 0.7,
         title = "Correlation Matrix — Rent Variables",
         mar = c(0,0,1,0))
```

#### Close Database Connection
```{r}
dbDisconnect(conn)
cat("Database connection closed.")
```

#### Regression Model: What Predicts Rent in the Twin Cities?
```{r}
# We want to answer: which factors predict higher rent?
# Our model: avg_rent ~ year + bedroom sizes

# First filter to complete rows only (no missing values)
df_model <- df %>%
  filter(!is.na(avg_rent) & !is.na(rent_1br))

cat("Rows used in model:", nrow(df_model), "\n")
```

#### Build the Regression Model
```{r}
# lm() = linear model
# avg_rent is what we are predicting (dependent variable)
# year, rent_1br, rent_2br, rent_3br are our predictors (independent variables)

model <- lm(avg_rent ~ year + rent_1br + rent_2br + rent_3br, 
            data = df_model)

# Show the full results
summary(model)
```

#### Interpret the Results
```{r}
# Extract key numbers from the model
r_squared  <- summary(model)$r.squared
adj_r2     <- summary(model)$adj.r.squared
f_stat     <- summary(model)$fstatistic[1]

cat("R-squared:         ", round(r_squared, 4), "\n")
cat("Adjusted R-squared:", round(adj_r2, 4), "\n")
cat("F-statistic:       ", round(f_stat, 2), "\n")
cat("\n")
cat("What this means:\n")
cat("  R-squared close to 1.0 = model explains rent variation well\n")
cat("  R-squared close to 0.0 = model explains very little\n")
```

#### Check Model Assumptions
```{r fig.height=8}
# These 4 plots check if our regression assumptions are met
# This is the same check you did in your MPG project!
par(mfrow = c(2, 2))   # arrange 4 plots in a 2x2 grid
plot(model)
par(mfrow = c(1, 1))   # reset back to normal
```

#### Predicted vs Actual Rent
```{r}
# Add predicted values to our data
df_model$predicted_rent <- predict(model)

# Plot predicted vs actual
ggplot(df_model, aes(x = predicted_rent, y = avg_rent)) +
  geom_point(color = "#185FA5", alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, 
              color = "red", linetype = "dashed") +
  labs(
    title = "Predicted vs Actual Rent",
    subtitle = "Points close to the red line = good predictions",
    x = "Predicted Rent ($)",
    y = "Actual Rent ($)"
  ) +
  theme_minimal()
```

#### Key Findings
```{r}
# Show the coefficients in a clean table
coef_table <- as.data.frame(summary(model)$coefficients)
coef_table <- round(coef_table, 4)
kable(coef_table, 
      caption = "Regression Coefficients",
      col.names = c("Estimate", "Std Error", "t Value", "p Value"))
```